---
title: "Borders"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
bibliography: Borders.bib
---

```{r setup, include=FALSE}
library("CBPData")
library("dplyr")
library("ggplot2")
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Specifically US Borders. Specifically, the US-Mexico Border.

US Customs and Border Patrol's reputation [leaves something to be desired](http://www.nytimes.com/2016/03/06/magazine/10-shots-across-the-border.html). The concerns about Border Partol are [not](http://www.nytimes.com/2015/10/15/us/aclu-accuses-border-patrol-of-underreporting-civil-rights-complaints.html) [new](http://www.nytimes.com/2014/05/06/us/complaints-of-abuse-by-border-agents-often-ignored-records-show.html).

Note on language: there's no legal ambiguity that crossing a national border without permission to do so, either explicitly (e.g. through a crossing point with a visa) or implicitly (e.g. between states within the European Union). It is thus factually correct to observe that people who cross a national border without permission to do so are in violation of the law. They have *illegally immigrated*. That said, the nation into which they've crossed does not assert that their *existence* is in violation of law, only their presence in that nation. Accordingly, I'm opting to avoid the title of "illegal immigrant" for the more neutral "undocumented immigrant". Assume for the remainder of this essay that references to "undocumented immigrants" actually means "people who entered the United States illegally." For a useful list of discussions on the topic, please see [Open Borders' Reference on the Topic](http://openborders.info/nomenclature-for-illegal-immigrants/).

# The Debate

## The Case for Open Borders

Labor Market Efficiency

Human Rights

## The Case for Closed Borders

If the governed want to exclude foreigners, it is their right to call for the government to do so, and the government's responsibility to respond to that call.

Economic Welfare: We enjoy a high standard of living because the government provides services that improve out lifestyles. However, larger populations reduce the government's ability to provide some of these services.

## Middle Grounds

The case for Open Borders with nations that have more progressive welfare systems than the US.

# Who Immigrates?

Why do people leave their homes to make the dangerous trek across the Southwest border into the United States?

## Economic Migrants

Why do people leave their homes to make the dangerous trek across the Southwest border into the United States?

Where are they coming from?

## Unaccompanied Alien Children and other Refugees

Why do children leave their homes to make the dangerous trek across the Southwest border into the United States?

Where are they coming from?

## "Bad Guys"

Drug runners, Alien Smuggling Organizations (ASOs) and their Coyotes

Why do people leave their homes to make the dangerous trek across the Southwest border into the United States?

Where are they coming from?

# Getting Here

The Train of Death

Alien Smuggling Organizations - Coyotes

# Getting In

The US-Mexico border is not heavily guarded between population centers. Border protections span from nothing to 18-foot-high walls.

How do they get in?

# Living in the US

How many undocumented immigrants are living in the United States?

How does living in the US affect their lifespan/health?

## Staying Under the Radar

Cash-for-Labor markets

# Getting Caught

How do they get caught?

What happens when they do? Detention Facilities

## Going Back

# Who Enforces the Border?

US Customs and Border Protection

Growing Budget:

```{r CBPBudget1}
nationFYear %>%
  filter(!is.na(Budget)) %>%
  ggplot(aes(x=FiscalYear, y=Budget)) +
  geom_line()
```

At the same time, the number of apprehensions has been declining:

```{r Apprehensions}
nationFYear %>%
  filter(!is.na(Budget)) %>%
  ggplot(aes(x = FiscalYear, y = Apprehensions)) +
  geom_line()
```

The combination of these factors is an agency whose effectiveness is inversely proportional to its budget.

```{r CBPBudget}
ggplot(nationFYear, aes(x = Budget, y = Apprehensions)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

And this degree of inefficiency is growing over time.

```{r}
nationFYear %>%
  mutate(CostPerApprehension = Budget/Apprehensions) %>%
  filter(!is.na(CostPerApprehension)) %>%
  ggplot(aes(x=FiscalYear, y=CostPerApprehension)) +
  geom_line()
```

# References
