---
title: "Fun with the Metaculus Leaderboard"
---


```{r}
library(rvest)
library(dplyr)
library(magrittr)
library(stringr)
library(lubridate)
library(plotly)
library(rmarkdown)
```

## Getting Started

To start, let's just scrape the current [all-time leaderboard](https://www.metaculus.com/rankings/#all).

```{r scrapeRanking}
rankingsTable <- read_html('https://www.metaculus.com/rankings/#all') %>%
  html_node('body > div > section > div > div > table')

rankings <- html_table(rankingsTable)

rankings %<>% mutate(
  profileURL = (rankingsTable %>% html_nodes('a') %>% html_attr('href')),
  playerNumber = as.integer(substr(profileURL, 19, 24))-100000
)

paged_table(rankings)
```

```{r}
plot_ly(rankings, x = ~Score, y = ~Rank, type = "scatter", text = ~Player)
```

```{r}
plot_ly(rankings, x = ~playerNumber, y = ~Score, type = "scatter", text = ~Player)
```

This is fun! If you divide the space into two equally-sized triangular halves, there's basically no one in the top-right half. If you think about it for a bit, it probably makes sense: Player number is a rough proxy for how long a player has had to make predictions. It takes a while to collect enough points from resolved points to make the leaderboard. Accordingly, there are no players with high player numbers (people who started very recently) and also high scores.

## Going a little deeper

Besides what's available in the raw leaderboard, there's a deeper layer of data available: individual profiles. It's a little rude to the server (since it's a hit on every profile in the leaderboard), but we can scrape these for some additional statistics, including number of questions forecast, number of resolved questions, and comments.

```{r}
rankings %<>%
  mutate(
    memberSince = NULL,
    predictions = 0,
    questionsPredicted = 0,
    resolutions = 0,
    comments = 0,
    questionsCommented = 0,
    questionsAuthored = 0
  )

for(i in 1:nrow(rankings)){
  profile <- read_html(paste0('https://www.metaculus.com', rankings$profileURL[i]))
  matches <- profile %>%
    html_node("body > div > div.content-wrap > section:nth-child(1) > div > div > div:nth-child(3) > span:nth-child(2)") %>%
    html_text() %>%
    mdy()
  rankings$memberSince[i] <- as.integer(matches[[1]][1])
  
  matches <- profile %>%
    html_node("body > div > div.content-wrap > section:nth-child(2) > div > div > div:nth-child(2) > span:nth-child(2)") %>%
    html_text() %>%
    str_extract_all('[0-9]+')
  rankings$predictions[i] <- as.integer(matches[[1]][1])
  rankings$questionsPredicted[i] <- as.integer(matches[[1]][2])
  rankings$resolutions[i] <- as.integer(matches[[1]][3])

  matches <- profile %>%
    html_node("body > div > div.content-wrap > section:nth-child(2) > div > div > div:nth-child(3) > span:nth-child(2)") %>%
    html_text() %>%
    str_extract_all('[0-9]+')
  rankings$comments[i] <- as.integer(matches[[1]][1])
  rankings$questionsCommented[i] <- as.integer(matches[[1]][2])
  
  matches <- profile %>%
    html_node("body > div > div.content-wrap > section:nth-child(2) > div > div > div:nth-child(5) > span:nth-child(2) > a") %>%
    html_text() %>%
    str_extract_all('[0-9]+')
  rankings$questionsAuthored[i] <- ifelse(is.na(as.integer(matches[[1]][1])),0,as.integer(matches[[1]][1]))
}

rankings %>%
  select(Player, memberSince:questionsAuthored) %>%
  paged_table()
```

It's conceivable to me that you can be bad a Metaculus for a long time, and still rise to the top of the leaderboard simply by virtue of having played longer. To avoid this, I'm more interested in a slightly different metric: average points-per-question.

```{r}
rankings %<>% 
  dplyr::mutate(pointsPerQuestion = Score/resolutions)

plot_ly(rankings, x = ~pointsPerQuestion, y = ~Rank, type = "scatter", text = ~Player)
```