---
title: "How to Price a House"
output:
  rmdformats::readthedown:
    css: site_libs/style.css
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For the [House Prices in Ames, Iowa](http://anthony.boyles.cc/portfolio/HousePricesInAmesIowa.html) competition, we looked at a ton of data on a ton of houses.

However, reality is much messier. This is a common complaint about Kaggle, and using Kaggle as a signal of experience doing data science. Kaggle provides its competitors with effectively perfect data, and then incentivizes really deep stacks of models to eek out tiny improvements in the competition's scoring metric.

So, when it came time to sell my own house, I might've liked to use all that nifty research I did for the Ames competition, but I couldn't because I don't have the same data from the town where my house is.

# Deciding a Listing Price

## Search Optimization

As potential buyers, we've mostly relied on Zillow and Redfin for information about houses we might want to purchase. These sites both have very nice query builders for filtering down possible houses.

However, there is one granular tool on top of a more continuous metric. Namely, they both have sliders to specify a desired price range, and the sliders move in increments of \$10,000. Moreover, as buyers, we're thinking in even more granular terms, rounding our intentions to buy to the nearest \$50,000. A good strategy, therefore, is to set a price that maximizes within that \$50,000 range. Unfortunately, that would recommend listing our house at something like \$449,900, which is less than we value the house, or \$499,000, which is more than we could possibly hope to get for it without extensive renovations.

## Competitive Optimization

At the time of listing, there weren't many houses similar to ours on the market yet. That meant we had some latitude to sell for a higher price relative to other houses like ours, because we aren't competing for buyers' attention.

These approaches kind-of work, but they really don't answer a fundamental question: what is the house worth? Rather, how much should we expect someone to pay for it?

# Predicting a Sale Price

All other things being equal (hah), selling a house is a trade-off between the amount of money you wish to gain from the sale and the amount of time you're willing to leave it on the market. If you're able to wait for someone willing to pay the price you want, the house can stay on the market for a long time. We are relatively flexible--we aren't trying to get rid of it quickly, but we'd like to minimize the time it's on the market and maximize the amount it sells for.

Zillow estimates that the ratio of sale prices to listing prices in our area is roughly .98. Stated differently, on average, a house near our house will sell for 98% of its list price.

Redfin publishes [historical aggregate data](https://www.redfin.com/blog/data-center) about real estate in a wide variety of different geographic areas. I've downloaded the [Redfin data](http://anthony.boyles.cc/data-cache/redfin.csv) for our town (Annandale, VA).

```{r, eval=FALSE}
library('readr')
library('magrittr')
library('dplyr')
library('lubridate')
library('plotly')

redfin <- read_csv("../data-cache/redfin.csv") %>%
  mutate(
    PeriodEnd = mdy(PeriodEnd),
    ListPrice = MedianSalePrice/AvgSaleToList
  ) %>%
  arrange(PeriodEnd)

plot_ly(redfin, x = ~PeriodEnd, y = ~ListPrice, type = "scatter", mode = "lines", name = "(Approx) Median List Price") %>%
  add_trace(y = ~MedianSalePrice, name = "Median Sale Price")
```

```{r, eval=FALSE}
library('stringi')
houses <- read_csv("../data-cache/redfin2.csv") %>%
  mutate(
    Sold = ifelse(`SALE TYPE` == "")
  ) %>%
  arrange(PeriodEnd)

plot_ly(redfin, x = ~PeriodEnd, y = ~ListPrice, type = "scatter", mode = "lines", name = "(Approx) Median List Price") %>%
  add_trace(y = ~MedianSalePrice, name = "Median Sale Price")
```
