---
title: "Inequality is Instrumental. Poverty is Terminal."
---

# Introduction

```{r setup, message=FALSE}
library("ShRoud")
library("magrittr")
library("dplyr")
library("intubate")
library("ineq")
library("plotly")
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

source("../infobox.R")
infobox(list("Epistemic Status"="Highly Uncertain", "Epistemic Activity"="Splitting Hairs"))
```

The US Consumer Financial Protection Bureau just won a [$180 million case against Wells Fargo](http://fortune.com/.../12/wells-fargo-cfpb-carrie-tolstedt/). The accusation was that Wells Fargo was engaging in unlawful account-opening practices, doing things like opening multiple accounts for consumers who only authorized the opening of one. This was a direct outgrowth of Wells Fargo's internal policies pressuring employees to meet certain account-opening quotas, or lose their jobs. Wells Fargo claims that it fired over 5300 employees over the last five years for unethically opening these accounts. In the meanwhile, Carrie Tolstedt, the executive who architected the employee incentive policy that led to this misbehavior retired with a \$126 million severance package.

I've spoken to people who are upset by the unceremonious firing of 5300 employees, but outraged by the severance bonus collected by Tolstedt. To me, this was exactly backwards: it's certainly in very bad taste for Tolstedt to collect such an obscene amount of money, but that in itself doesn't hurt anyone's livelihood. Throwing 5300 households into financial turmoil does hurt people's livelihood, and that is what outrages me.

Perhaps inequality is a red herring. Perhaps we should talk about fixing poverty.

# Inequality and Poverty

Imagine a world in which there are 100 people. The 11 poorest people live in terrible conditions, eating only 1200 calories per day on average. By contrast, the wealthiest person is *obscenely* wealthy, controlling more wealth than the 90 poorest people. This is a world with a very high degree of inequality. It looks something like this:

```{r}
base = 1.06

data <- data.frame(
    x = 1:100,
    y = base^(1:100)/base^100
)

data %>%
  plot_ly(x=~x, y=~y) %>%
  add_bars()
```

Where each bar shows the amount of wealth held by the xth wealthiest person *and* everyone poorer than them. This is pretty close to the world we actually live in. In contrast, let's consider a world in which wealth is uniformly distributed. It looks like this:

```{r}
data.frame(
    x = 1:100,
    y = 1:100/100
  ) %>%
plot_ly(x=~x, y=~y) %>%
  add_bars()
```

In other words, the second graph is something like an ideal communist society. If we understand these types of graphs, they can tell us a lot about the distribution of wealth in a society. However, they don't give us a simple metric by which to measure inequality. For that, we have the [Gini Coefficient](https://en.wikipedia.org/wiki/Gini_coefficient). The Gini coefficient of the first graph is `r formatC(ineq(base^(1:100)), digits = 2)`. The Gini coefficient of the second graph is 0.

There are, broadly speaking, two strategies to reducing inequality. You can 1) increase the wealth of the poor people, or 2) decrease the wealth of the rich person. Since it's easier, let's focus on strategy 2 for a moment.

This is a fun strategy. We get to blame someone who is well-off for all of our problems! We probably won't even be wrong! So, we grab the rich person, drag them to court and tell the government all about all the horrible things they've done. They get slapped with crushing fines on top of crushing legal bills, and they aren't so rich anymore. We all go home, delightedly thinking "justice has been served."

The only problem is what happens when we get home. The 11 poorest people are still poor, hungry and miserable. The easy strategy for solving inequality didn't help them at all. Their problem was never inequality. It was poverty.

> ... What? No one is suggesting just destroying the rich person's wealth. The point is that you take some of that wealth and GIVE IT TO THOSE STARVING PEOPLE.

I *know* no one is proposing burning money.

<iframe src="https://player.vimeo.com/video/175819562#t=259s" class="center-block" width="640" height="360" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen></iframe>

Well, *I'm* not, anyway.

Let's go back and focus on inequality-reducing strategy 1 (increase the wealth of poor people). There are a lot of ways we can do that. One is to take money from the wealthy by taxation, punitive fines, and social pressure to effective philanthropy, and give it to the poor. However, there may be other viable strategies by which poverty can be reduced. These may not necessarily have a huge effect on inequality.

For example, let's consider a perfectly unequal society. No one owns anything, except for one person who owns everything.

```{r}
data.frame(
    x = 1:100,
    y = c(rep(0, 99), 1)
  ) %>%
plot_ly(x=~x, y=~y) %>%
  add_bars()
```

This is a distopian hellscape in which 99 people scratch a living off of rocks, necessarily stealing from the only person with anything to steal. Its Gini Coefficient is `r ineq(c(rep(0, 99), 1))`.

Now, instead of that world, let's say that the bottom 99 people each hold \$25,000 in wealth, and the [top person holds \$75 billion](http://www.forbes.com/billionaires/list/). Like this:

```{r}
data.frame(
    x = 1:100,
    y = c(1:99*25000, 7.5e10)/7.5e10
  ) %>%
plot_ly(x=~x, y=~y) %>%
  add_bars()
```

Let's further imagine that [the poverty line is around \$1.90 per day](https://en.wikipedia.org/wiki/Poverty_threshold). An investment of $25,000 compunded anually at 3%, yields about \$750 per year, or \$2.05 per day, pushing everyone over that threshold. This is also a staggeringly unequal society, with a Gini coefficient of `r formatC(ineq(c(cumsum(rep(25000, 99)), 7.5e10)), digits=2)`. However, the quality of life enjoyed by the bottom 99 is higher than the quality of life enjoyed by the poorest people today. In other words, poverty can **theoretically** be solved more-or-less independently of inequality.

Back in the real world, there are 7.3 Billion people today (and that number is [all but certain to grow for awhile](https://esa.un.org/unpd/wpp/Publications/Files/Key_Findings_WPP_2015.pdf)). Endowing each one with an investment account of \$25,000 (as in the thought experiment above) will cost around \$182.5 Trillion. For comparison, the current approximate [value of everything](https://publications.credit-suisse.com/tasks/render/file/?fileID=BCDB1364-A105-0560-1332EC9100FF5C83) is $241 Trillion, meaning that more than 75% of the assets in the world would need to be reallocated to accommodate such a scheme. Don't get me wrong--it's a great way to solve poverty. It's just not going to work.

So here's the question: if the way to reduce poverty is to transfer wealth to the poorest, where are we going to get that wealth?

# Where does the money come from?

> Unless you punish schemes to transfer wealth upward via fraud, which are fairly ubiquitous, you won't be able to outpace the rate at which your attempts to cure poverty are siphoned up by the rentiers.

Some industries ARE highly effective at siphoning money upwards (high risk credit markets, for instance). But whenever I start thinking about these types of markets, I bump into a problem. This is a market niche because there is a real and urgent need for it. If we eliminate payday and auto title lenders, what will poor people who rely upon these institutions do? There are some hopeful answers to this question ([Lending Clubs](https://www.lendingclub.com/), for example), but they haven't been able to supplant the lecherous practices against which they're competing.

But back to the question at hand: does this all mean that inequality is a more important problem than poverty? I think not.

# Inequality as a Terminal Value

Inequality itself [may be a bad thing](http://www.imf.org/external/pubs/ft/wp/2016/wp16121.pdf). If we consider the health of the economy as a driver of the welfare of everyone (except, arguably, the excessively rich), then a healthy economy is helpful (and potentially vital) to the cause of global welfare improvement. The distribution of wealth may "cause" considerable externalities many key parts of the economy, including net spending, demand for labor, and thus net employment availability, culminating in income for the median person.

However, this isn't the entire picture. This, so far as I can tell, is contingent upon viewing the economy as a zero-sum game. The economy is slow because wealthy people have wealth and they don't spend it as rapidly as poorer people (everyone else) does. While the latter statement is empirically true, the former is false. It is possible to *create* value, as well as to take it. Some industries are very good at creating value; technology, for example.

Imagine a world in which all practical avenues to upward money siphons has been sufficiently restricted, so that it's easier to make money using comparatively honest means (selling useful things at market prices, for example). That doesn't obviate the world of its Bill Gateses and Mark Zuckerburgs. They are wealthy because they made things that other people would then buy for billions of dollars.

Now Imagine a world in which everyone possessed sufficient income to ensure their access to safe housing and adequate nutrition (*cough* UBI *cough*), that also doesn't obviate us of the super-rich. It does, however, eliminate some of the worst burdens of deep poverty. This could continue to be a world with deep inequality. But its problems wouldn't share the same level of moral importance as that of the world we live in today.

There may be a version of this argument (highly unequal wealth distributions lead to anemic macroeconomic conditions) that can convince me that inequality holds primacy over poverty in importance, but my threshold is going to be pretty high. In order for macroeconomic distortions resulting from inequality to be more urgent than poverty, those distortions must reach all the way down to the quality of life of the impoverished. I can imagine economic fictions in which this could happen, but I'm bounded by only a passing familiarity with macroeconomics. Even so, they don't seem especially plausible to me.

# Poverty Reduction is the thing we should value

Poverty-mitigation strategies are likely to have measurable effects on inequality. So why quibble over which is more important?

