---
title: "Longevity"
bibliography: Longevity.bib
---

# Introduction

```{r, message=FALSE}
library("htmltab")
library("magrittr")
library("plyr")
library("tidyverse")
library("plotly")
library("intubate")
library("knitr")
library("htmltab")
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
```

```{r, results='asis'}
source("infobox.R")
infobox(list(
  "Status" = "Incomplete",
  "Epistemic Status" = "Speculative",
  "Fork of" = tags$a(href='http://www.gwern.net/Longevity', "Gwern's Longevity"),
  "First Revision" = "2016-10-14",
  "Last Revision" = as.Date(file.info("drafts/Longevity.Rmd")$mtime)
))
```

I want to live for a very long time.

There are, broadly speaking, two ways to extend lives. One is public health. To extend lives using public health, go to an actuarial table and ask "Which of these causes of death can be mitigated?" Then pick one, like car accidents and figure out how to stop people from dying in car accidents. Invent seatbelts and airbags, and then convince governments to make laws forcing people to use seatbelts and airbags. And then people start dying less from car accidents, and you've extended their lives.

The people doing public health are really, **really** good at it. [Average human lifespans are increasing all over the world](https://ourworldindata.org/life-expectancy/#life-expectancy-has-improved-globally), almost entirely due to victories in public health. People are [getting vaccinated](https://data.oecd.org/healthcare/child-vaccination-rates.htm). Eating better food. Breathing less of fewer toxic things. Drinking cleaner water. Breastfeeding. [Hurting each other less](https://en.wikipedia.org/wiki/The_Better_Angels_of_Our_Nature).

Basically, public health is awesome, and we should be handing out WAY more money and medals to people doing public health. However, interventions of the type advocated by the public health community generally (so far) can only get people as far as old age [@Olshansky1990-vd]. After a point, seatbelts and clean water aren't going to make a huge difference, because you're probably going to die of a natural cause like heart failure or some sort of cancer [@Werfel2015-zx].

So, the other way to extend lives is to stop people from dying of natural causes. We don't know very well how to make that happen, but we have [a research agenda](https://en.wikipedia.org/wiki/Strategies_for_Engineered_Negligible_Senescence) that (we hope) can figure it out [@De_Grey2007-ac]. Moreover, there are a lot of people who got rich, decided they want to live for a really long time too, and set up [well-funded companies](https://www.calicolabs.com/) and [non-profits](http://glennfoundation.org/) to search for other avenues to life extension. In the meanwhile, there are a small number of odd things that we have reasonable evidence extends lifespans.

After more than a century of concerted effort and tinkering with a bewildering variety of thousands of interventions, it can safely be said that almost all attempted interventions have failed, so the prior odds are strongly against any new one working. Even so, for those of us who want to give ourselves a fighting shot at extreme longevity, even tiny longevity advantages may be worthwhile. That's because we don't know how soon the big breakthrough will come that enables our oldest people to roar through their early 120's. If you're alive when the breathrough does happen, that gives you a non-zero probability of benefitting from it, and living a long time yourself. Alternately, if you die before the big breakthrough, you're almost certainly not going to benefit from it.

So it's worth exploring whether certain interventions will result in a net improvement in your probability of living long enough to live indefinitely. However, this probability is likely to come at a non-trivial cost. As [Clement Freud once said](https://en.wikiquote.org/wiki/Clement_Freud#Sourced), "If you resolve to give up smoking, drinking and loving, you don't actually live longer; it just seems longer." (see also: ["Which Risks of Dying Are Worth Taking?"](http://www.spencergreenberg.com/2013/10/which-risks-of-dying-are-worth-taking/))

What I want to figure out is how to reduce all-cause mortality (ACM) as estimated in randomized trials on healthy adult human populations for which there are enough trials to do a random-effects/multilevel model, so I can estimate a posterior predictive interval of the benefit of an intervention, letting me compare mortality reduction with intervention cost, from which I can compute an expected value. Whew, that's a head-full. Let's break out those constraints individually:

**1. reduce "all-cause mortality",**

ACM is the end-point of choice for life-extension because it is both closest to what we care about (additional life) and it is the hardest of all end-points - it is difficult to cheat, miscount, or not notice, and by being simple, consistent, and unambiguous, avoids many failure modes such as subgroup hacking or dropping outliers or tweaking covariates (either a larger fraction of the controls are dead or not). It can also typically be extracted from most studies, even if they otherwise fail to report much of the data.

Average reduction in all-cause mortality is also excellent for ignoring any zero-sum tradeoffs: it's no good if reported benefits are only because the doctors writing up a trial kept slicing deaths by different categorizations of causes until they finally found a _p_<0.05 or if the gains only appear by excluding some deaths as 'irrelevant' or if the gains in one cause of death comes at the expense of another cause (is it really helpful to avoid death by heart attack if the intervention just means you die of cancer instead?).

From a life-extension perspective, it is irrelevant if an intervention reduces some kinds of diseases while making one much more prone to dying from routine causes such as injuries or minor operations under anesthesia (especially since, in modern societies, one is all-but certain to undergo some surgery at some point).

In short, an intervention either reduces all-cause mortality or it doesn't.

**2. as estimated in randomized trials,**

[Correlation is not causation]() and correlates rarely turn out to be causal. With aging, the situation is especially severe since aging is the exponential increase in mortality with time as all bodily systems gradually begin to fail; one cannot simply take a correlation of some chemical with aging or mortality and expect an intervention on it to have more than a minute chance of making a difference, since most (if not nearly every) chemical or substance or biomarker will change with age.

The best way to insulate our conclusions from selection effects is to perform large, controlled experiments, rather than to rely upon natural data.

**3. on healthy adult human populations**

Studies of interventions in the sick are unhelpful, as any observable benefits from such an intervention hopefully come from the particular disorder being treated. One special challenge for this criteria is studies done in elderly populations. Because death rates amongst the elderly are so high, they offer higher power to detect reductions in mortality. However, that death rate is high because of the aging process itself, and so subjects will tend to have many chronic conditions and diseases as well. These confound otherwise useful results, and thus must be treated with some caution.

Animal studies are also unhelpful as [results tend to be unreliable across species](). One particularly noteworthy source of unreliability is that some species age in entirely different ways than humans. Perhaps it is better to say that humans naturally live unusually long lives for animals, and thus age in very different ways than other organisms.

**4. for which there are enough trials to do a random-effects/multilevel model, so we can extract a posterior predictive interval of the benefit,**

It's important to include heterogeneity from population to population as part of the uncertainty. One of the subtleties often missed in dealing with modeling is that a 95% CI around a parameter is not a CI around the outcome or result.

**5. letting us compare mortality reduction against intervention cost,**

For making decisions, the costs of the intervention must be taken into account. A taxing exercise regimen may be proven to increase longevity, but is that of any value of the regimen takes away much more of your life than it gives back? It may be that no matter whether our certainty is 99% or 99.99% that it works, we would never choose to embark on that regimen. On the other hand, for a cheap enough intervention, we may be willing to accept substantially lower probability: baby aspirin, for example, costs \$7 & <10 seconds each morning / <60 minutes a year and we might be willing to use it if our final probability of any benefit is a mere 50%.

**6. then yielding an expected value of the intervention (and if <\$0, perhaps further analysis about whether it has a reasonable chance of ever becoming profitable and what the Value of Information might be of further trials)**

If an intervention passes all these criteria, we can be highly confident that we will gain life, rather than lose it.

Of course, these criteria are stringent: we can, for example, exclude almost everything which might appear in an issue of Life Extension Magazine for being too small, conducted on animals or sick people, not being a randomized experiment, or reporting benefits only on an extremely specific biological endpoint like some cholesterol-related metric you've never heard of. Medical journals don't offer rich pickings either, as they understandably tend to focus on experiments done in sick people rather than healthy people (apparently there's more funding for the former - who knew?). And when a randomized experiment _is_ done in healthy adult humans, the experiment may either be too small or have run for too short a time, since annual mortality rates for many participant groups are 1% or less; for any sort of effect to show up, you want (very expensive, very rare) studies which enroll at least a thousand subjects for a decade or more. It would not be surprising if we must intone the ritual ending to all Cochrane Reviews, "there is insufficient evidence for a recommendation".

Unfortunately, there are precious few interventions which meet these stringent requirements. Accordingly, I'll be looking at many interventions for which there is **some** evidence which does not meet this rigorous set of expectations.

## Interventions

Evaluated Supplements/Medications:

* [Vitamin D3](#vitamin-d)
* [Metformin](#metformin)

Possible Supplements/Medications:

* [Aspirin](https://www.google.com/search?q=caffeine+all+cause+mortality)
* [Rapamycin](http://en.wikipedia.org/wiki/rapamycin)
* [Nicotinamide Riboside]()
* [Resveratrol]()
* [Pterostilbene]()
* [Bisphosphonate](http://en.wikipedia.org/wiki/Bisphosphonate)
* [D-Glucosamine](http://www.ncbi.nlm.nih.gov/pubmed/?term=22828954)
* [Magnesium](http://www.ncbi.nlm.nih.gov/pubmed/?term=24259558)
* [Lithium Chloride](http://www.ncbi.nlm.nih.gov/pubmed/?term=21301855)
* [Omega 3]()
* [Acetyl L-Carnitine]()
* [Phosphatidylcholine]()

Possible Dietary Modifications:

* [Mediterranean Diet](http://www.nejm.org/action/showImage?doi=10.1056%2FNEJMoa1200303&iid=f01)
* [Caloric Restriction](http://en.wikipedia.org/wiki/caloric%20restriction)
* [Coffee/Caffeine](http://www.nejm.org/doi/full/10.1056/NEJMoa1112010#t=article)
* [Alcohol](http://en.wikipedia.org/wiki/Long-term%20effects%20of%20alcohol%20consumption)
* [Nuts]()

Possible Lifestyle Modifications:

* [Aerobic Exercise]()
* [Resistance Exercise]()
* [Accident Risk Reduction]()
* [Castration](http://en.wikipedia.org/wiki/Castration%23Medical%20consequences)

In the exceedingly unlikely case that I exhaust all of these, there's the...

* [Geroprotectors Database](http://geroprotectors.org/)
* [DrugAge](http://genomics.senescence.info/drugs/browse.php?search=)

## Definitions

### Value of Life

The valuation of one year of life I am using here [is \$50,000](http://en.wikipedia.org/wiki/Value%20of%20life%23Life%20Value%20in%20the%20US "Wikipedia: Value of life#Life Value in the US"), due to its commonness and conservativeness (some valuations are much higher, and so favor the use of interventions with much lower certainty thresholds).

```{r valueOfDALY}
#If you'd like to see how differently valuing your life-years changes the outcomes, just change this variable:
valueOfDALY <- 50000
```

### Population survival curve

To work with life expectancies and changes in it, let's compute a standard [Gompertz curve](http://en.wikipedia.org/wiki/Gompertz%20curve) model of age-related mortality, using a the [US Social Security Administration's 2013 Actuarial Tables](https://www.ssa.gov/oact/STATS/table4c6.html).

(For an excellent introduction to the Gompertz curve, see [this blog post](https://gravityandlevity.wordpress.com/2009/07/08/your-body-wasnt-built-to-last-a-lesson-from-human-mortality-rates/).)

```{r ActuarialTable}
if(!file.exists('../data-cache/actuarial.csv')){
  temp <- htmltab('https://www.ssa.gov/oact/STATS/table4c6.html', which=2, headerSep = ' ')
  names(temp) <- c('Age', 'MaleDeathProbability', 'MaleNumberOfLives', 'MaleLifeExpectancy', 'FemaleDeathProbability', 'FemaleNumberOfLives', 'FemaleLifeExpectancy')
  temp %>%
    filter(row_number() < 121) %>%
    write_csv('../data-cache/actuarial.csv')
}
gendered_actuarial <- htmltab("https://www.ssa.gov/oact/STATS/table4c6.html", which = 2, headerSep = " ") %>%
  slice(1:(n()-1)) %>%
  transmute(
    Age = as.numeric(`Exact age`),
    extantMen = as.numeric(gsub(',', '', `Male Number of     lives`)),
    extantWomen = as.numeric(gsub(',', '', `Female Number of     lives`))
  )

plot_ly(gendered_actuarial, type = 'scatter', mode = 'lines') %>%
  add_lines(x = ~Age, y = ~extantWomen/100000, name = "Women") %>%
  add_lines(x = ~Age, y = ~extantMen/100000, name = "Men")
```

From this, we can see that women at every age have higher survival rates than men of the same age. However, rather than stratify everythin on gender, I'm just going to combine the two surviving groups. It's a blunt approach, but it's hardly the worst available option. Let's compare this to an estimate of survival fractions of Dutch adults [@Cramer2013-xo].

```{r GompertzCurve}
S <- function(t, RR=1) exp(-(RR*0.000016443)/log(1.1124) * (1.1124^t - 1))

actuarial <- gendered_actuarial %>%
  dplyr::group_by(Age) %>%
  dplyr::summarise(American = (extantMen + extantWomen)/200000) %>%
  dplyr::mutate(Dutch = S(Age))

plot_ly(actuarial, x = ~Age) %>%
  add_trace(y = ~Dutch, name = "Dutch", mode = 'lines') %>%
  add_trace(y = ~American, name = "American", mode = 'lines') %>%
  layout(yaxis=list(title='Surviving Proportion'))
```

Here we can see that Dutch people are less likely to die throughout early and middle age, but Americans are less likely to die later in life. This reversal happens sometime around age 67.

```{r GompertzSimulation}
## the `min` is because past age 104, hazard>1 which is wrong, so need a ceiling
H <- function(t, RR) min(1, (RR*0.000016443)*1.1124^t)

## for each possible age, randomly sample a life event according to that age's hazard;
## then the subject dies at the age at which their first FALSE is found.
## So `TRUE,TRUE,TRUE,FALSE,TRUE,FALSE` ~> died 4yo.
rgompertz <- function(n, RR=1) {
  ageHazards <- sapply(0:120, function(t) H(t, RR))
  replicate(n, Position(`!`, sapply(ageHazards, function(h) sample(c(TRUE,FALSE), 1, prob=c(1-h, h)))))
}
```

### Converting risk reduction to average life expectancy gain

A naive interpretation of a risk reduction (RR) is to treat it as though it can "restore" some number of years of life. Consider a someone with a life expectancy of around 80 years. Now let us suppose this person decides to start a life-extending activity with an RR of .85. The gain we might hope to get from this activity is $(1-.85)*80=12 \text{years}$. This is, sadly, strongly optimistically biased.

The Gompertz hazard curve can be multiplied by the RR to give a particular trajectory with that change in risk, and the mean can be obtained by integrating over a lifetime. Doing two such integrations, one with a baseline RR of 1 and one with the new RR, then gives the two different life expectancies, and their difference indicates the net gain/loss:

```{r lifeExpectancy}
lifeExpectancy <- function(RR, age1, age2) {
  integrate(function(t) S(t, RR) / S(age1, RR), age1, age2)$value
}

lifeExpectancyGain <- function(RR, startingAge=0, endingAge=Inf){
  lifeExpectancy(RR, startingAge, endingAge) -
  lifeExpectancy(1, startingAge, endingAge)
}

rrs <- seq(0.01, 2, by=0.01)
gainsVsRR <- data.frame(RR=rrs, Gain=sapply(rrs, lifeExpectancyGain))

plot_ly(gainsVsRR, x=~RR, y=~Gain, type = 'scatter', mode = 'lines')

# Side note: it looks like a logarithmic curve. If we needed better computational performance, we could exploit this property to replace the integrals with cheaper arithmetic operations:

gainlm <- gainsVsRR %>%
  ntbt_lm(Gain ~ log(RR))

lifeExpectancyGainApproximation <- function(RR) predict(gainlm, data.frame(RR=RR))
```

From this, we can see a much more realistic relationship between RR and life years gained. An RR of .85 delivers around ~1.6 years.

There is a fortunate upside to this phenomenon. If we wait to use an intervention until what might seem like 'late' in life, we will not have suffered a large opportunity cost (of ~40%) like one might think, but much less (~1.7%). The gains are minimal early on, and the opportunity cost only begins to really mount starting in one's 50s or 60s, which has the important implication that such interventions will benefit younger people much less (at least when it comes to interventions which offer only a constant reduction in risk and do not affect the acceleration of mortality caused by the aging process itself) and so the risks are more likely to make interventions profitable.

### Maximum possible profit

More generally, it would be interesting to get an idea of what is the most profit possible for any particular reduction in ACM, by granting the most favorable possible assumptions: certainty that the RR is exact, that the intervention is side-effect-less and free both to start and continue, and it can begin at birth if need be. For the plausible range of RRs, 0.5-1.0:

```{r profitByAge}
profitByAge <- function(t, RR=0.85, yearValue=valueOfDALY, annualCost, startCost=0, probabilityPenalty=(1/3)) {
  (lifeExpectancyGain(RR, startingAge=t) * yearValue * probabilityPenalty) -
  (lifeExpectancy(RR,t,Inf) * annualCost) + 
  startCost
}

rrs <- seq(0.5, 1, by=0.025)

startAge <- 0
data.frame(
  RR = rrs,
  Years         = sapply(rrs, function(r) lifeExpectancyGain(r, startAge)),
  MaximumProfit = sapply(rrs, function(r) profitByAge(startAge, RR=r, annualCost=0, startCost=0, probabilityPenalty=1))
) %>%
  kable()
```

This provides upper bounds, and can help gauge plausibility of interventions for various variables - for example, if we were a 90-yo man considering embarking on caloric restriction, we might note that the upper bounds look like:

```{r startingAt90}
rrs <- seq(0.8, 1, by=0.05)
startAge <- 90
data.frame(
  RR = rrs,
  Years         = sapply(rrs, function(r) lifeExpectancyGain(r, startAge)),
  MaximumProfit = sapply(rrs, function(r) profitByAge(startAge, RR=r, annualCost=0, startCost=0, probabilityPenalty=1))
) %>%
  kable()
```

Then we have to take into account the uncertainty that caloric restriction works at all in the extremely elderly (a halving is probably optimistic), the high probability its reduction in mortality will be relatively modest and in the 0.90s, the large financial & time costs of safely doing such a stringent diet (easily thousands of dollars in time and higher-quality food ingredients), the need to learn new recipes and count calories, the possible side-effects like increasing [frailty](http://en.wikipedia.org/wiki/Frailty%20syndrome "Wikipedia: Frailty syndrome") & [muscle loss](http://en.wikipedia.org/wiki/Sarcopenia "Wikipedia: Sarcopenia"). Given that the upper bound for plausible effect sizes is merely 1 year/\$50k, it seems inadvisable for such a person to take up a Caloric Restriction Diet.

Using this approach, we can also reverse the question. Given an RR and a set of costs, we can determine what the optimal age to begin a regimen of life-extension activities.

```{r}
costToRR <- function (startingAge=30, annualCost, startCost=0, probabilityPenalty=1) {
  uniroot(f=function(r) {
    profitByAge(startingAge, RR=r, annualCost=annualCost, startCost=startCost,
                probabilityPenalty=probabilityPenalty)
  }, lower=.Machine$double.eps, upper=1)$root
}
```

### Power analysis

How many subjects does it take to detect a specified RR? This is useful for planning clinical trials but also in interpreting them. We can do simulation power analyses by using `rgompertz` to generate a number of subjects' death dates for different RRs (eg 0.85 vs 1.0), counting how many deaths in each group fall within a specified followup period of years, and running some sort of analysis based on that data.

```{r gompertzSim}
gompertzSimulation <- function(RR, n1, n2, startingAge, followupYears) {
  experimental       <- Filter(function(deathDate){deathDate>=startingAge}, rgompertz(n1, RR))
  experimentalDeaths <- Filter(function(deathDate){deathDate<=(startingAge+followupYears)}, experimental)

  control            <- Filter(function(deathDate){deathDate>=startingAge}, rgompertz(n2, 1))
  controlDeaths      <- Filter(function(deathDate){deathDate<=(startingAge+followupYears)}, control)

  df <- data.frame(RR=RR, Start=startingAge, Followup=followupYears,
                   N1=length(experimental), N2=length(control),
                   N1.deaths=length(experimentalDeaths), N2.deaths=length(controlDeaths))
  return(cbind(df, data.frame(RR.observed=(df$N1.deaths/df$N1) /(df$N2.deaths/df$N2))))
}

gompertzPower <- function(RR, n1, n2, startingAge, followupYears, iters=100) {
  1:iters %>%
    mclapply(function(i) gompertzSimulation(RR, n1, n2, startingAge, followupYears)) %>%
    ldply()
}
```

Given a function to estimate power for a particular set of trial characteristics, we can also search for the sample size which would make a trial (or fixed-effect meta-analysis) of particular RR effects well-powered as a heuristic for what sort of sample sizes we should expect (remembering that being well-powered is neither necessary nor optimal for decision-making):

```{r, eval=FALSE}
#This is turned off because it takes a really long time to run.
gompertzPowerSearch <- function(RR, startingAge=50, followupYears=10, targetPower=0.8, startingN=4000, iters=100, increment=200){
  n <- startingN
  nPower <- 0
  while (nPower < targetPower){
    n <- n+increment
    sims <- gompertzPower(RR, n, n, startingAge, followupYears, iters)
    pvalues <- sapply(1:nrow(sims), function(i) {
      prop.test(
        c(sims[i,]$N1.deaths, sims[i,]$N2.deaths),
        c(sims[i,]$N1, sims[i,]$N2),
      alternative="less")$p.value
    })
    nPower <- sum(pvalues<=0.05) / length(pvalues)
  }
  return(n)
}

startingN <- 4500

for (RR in seq(0.80, 0.95, by=0.01)) {
  n <- gompertzPowerSearch(RR, startingN = startingN)
  startingN <- n
  print(paste0("RR ", RR, "; necessary sample: ", n))
}
```

## Prior on RRs

Useful interventions are rare and our analyses should incorporate a skeptical informative prior reflecting our knowledge that it is highly unlikely that any particular intervention will genuinely increase life expectancy; in particular, RRs <0.5 (or >2.0) are rare. Since there are no previous compilations of randomized effects for interventions on all-cause mortality in healthy people to use for priors, I fall back on general-purpose but still informative skeptical priors found to work well across many problems by [Gelman et al 2008](http://projecteuclid.org/download/pdfview_1/euclid.aoas/1231424214 "A weakly informative default prior distribution for logistic and other regression models") & [Pedroza et al 2015](http://gwern.net/doc/statistics/2015-pedroza.pdf "Performance of informative priors skeptical of large treatment effects in clinical trials: A simulation study"): the Cauchy(0, 2.5^2) and Normal(0, 0.35^2) priors.

# Nicotinamide Riboside

The current supplement *du jour* for life-extension enthusiasts is called [Basis](https://www.elysiumhealth.com/basis) by Elysium Health. Its backing is unusually good, but it hasn't been subjected to any clinical testing (intentionally opting to go to market as a supplement in an attempt to side-step FDA regulation, which would take decades of testing and billions of dollars). Based on its list of scientific advisors, it may have sufficient theoretical support to be worth evaluating.

It's composed of Nicotinamide Riboside (250mg/pill) and Pterostilbene (50mg/pill). However, they claim that the [current batch's doses](https://www.elysiumhealth.com/purity-guide) are more like 277mg/pill and 54mg/pill respectively.

It's interesting to me in part because it is comparatively expensive. Elysium Health is selling a subscription model in which Basis may be purchased annually (\$480), semi-annually (\$45), monthly (\$50), or a la carte (\$60). These prices are, [surprisingly](https://smile.amazon.com/s/ref=nb_sb_noss?url=search-alias%3Daps&field-keywords=Nicotinamide+Riboside) [competitive](https://smile.amazon.com/s/ref=nb_sb_noss?url=search-alias%3Daps&field-keywords=Pterostilbene) with other supplement providers offering the same supplements at similar dosages.

# Pterostilbene

# Rapamycin

Rapamycin is famous for life-extension results in animal experiments (especially mice), but the animal experiment environments may be unrealistic (raised under controlled, sometimes [Specific Pathogen Free](http://en.wikipedia.org/wiki/Specific-pathogen-free) conditions) and rapamycin's primary use in humans is as an immunosuppressant and has been linked with bad outcomes. Optimal dosage is unclear; at least one mice study found a higher dose did not extend lifespan as much while increasing certain cancer rates ([Bitto et al 2016](https://elifesciences.org/content/5/e16351)). Rapamycin may never undergo clinical trials in healthy adult humans because the risks are too great, although animal trials such as [a large-dog trial](http://www.nytimes.com/2016/05/17/us/aging-research-disease-dogs.html), continue.

# Bisphosphonates

* https://www.fightaging.org/archives/2011/02/bisphosphonates-and-an-unusual-longevity.php
* https://www.fightaging.org/archives/2015/12/more-investigation-of-bisphosphonates-and-reduced-mortality.php

# Alcohol

"A glass a day" has long been noted to correlate with longer life anywhere one looks, but the confounds for moderate responsible drinking are too noticeable to put much faith in it and the debate continues to rage. [It's troubling](http://davidroodman.com/blog/2015/05/01/are-the-benefits-of-moderate-drinking-a-myth/) that the economics literature seems to find health benefits to taxing alcohol, and that the more powerful [Mendelian randomization](http://en.wikipedia.org/wiki/Mendelian%20randomization) design (while far from infallible) shows harm rather than benefit. But the real issue is that despite seeming like the easiest experiment ever to run (hand out free booze), there apparently are no randomized trials.

# Caloric Restriction

Caloric Restriction is promising, but current randomized studies are too short to show any effect on mortality (e.g. in [Ravussin et al 2015](http://biomedgerontology.oxfordjournals.org/content/70/9/1097.full). For example, no one died, so there is no all-cause mortality estimate.

For example, one primate study of caloric restriction found benefits only if it [excluded a bunch of deaths in the CR group](http://junkfoodscience.blogspot.com/2009/07/calorie-restrictive-eating-for-longer.html) and defined them as not age-related, ignoring "monkeys who died while taking blood samples under anesthesia, from injuries or from infections, such as gastritis and endometriosis." (Even in the [followup paper](http://www.nature.com/ncomms/2014/140401/ncomms4557/full/ncomms4557.html "'Caloric restriction reduces age-related and all-cause mortality in rhesus monkeys', Colman et al 2015") several years later, the all-cause mortality reduction is much smaller than the "age-related" subcategory.)

# Castration

If you read that heading twice, Yeah, it means exactly what you think it means. Castration is [often reported](http://lesswrong.com/lw/lm4/effects_of_castration_on_the_life_expectancy_of/) to increase longevity. However, we can't be too certain about these reports. For one thing, studies of historical groups of eunuchs are inconsistent (e.g. PMID: 8232579) and often based on small sample sizes (n < 100). Studies of dogs show [no large effect of castration](http://www.straightdope.com/columns/read/3068/does-castration-longer-life).

The best sort of study might be to do a Mendelian randomization for testosterone levels on a large cohort and see whether a lower genetic load for testosterone affects lifespan (adjusting for other genetic factors). This might be possible to do in the UK Biobank data, as testosterone MR studies have been done before (though I haven't looked in detail into these studies). http://www.nature.com/articles/srep21306

But we could offhandedly reject it as impractical for a number of reasons: too serious a reduction in quality of life ([going well beyond the loss of sex](http://en.wikipedia.org/wiki/Castration%23Medical%20consequences)), social ostracism, probably must be done before puberty for any benefits, may not be legal, not verified in any randomized trials (a concern because not all historical eunuch seem to show life expectancy gains; [Egyptian eunuchs reportedly](http://lesswrong.com/lw/lm4/effects_of_castration_on_the_life_expectancy_of/cogu) had severe losses), and at least one of the proposed mechanisms may have lost efficacy in the modern setting (greater resistance to infectious diseases).

Moreover, there may be stronger hypotheses in measuring estrogen, or estrogen/testosterone ratios, or some other similar metric. Female mice tend to outlive male mice. However, [treating male mice with estrogen leads to substantially longer lifespans](http://onlinelibrary.wiley.com/doi/10.1111/acel.12496/full) (a stronger effect than metformin). Solid research suggests that estrogen treatment in mice promotes stem cell self-renewal, which is a clear mechanism for how estrogen could promote lifespan (and explains its effects on multiple organ systems, including the immune system): http://www.nature.com/nature/journal/v505/n7484/full/nature12932.html

# Coffee

Being the kind of person who drinks coffee is really bad for you. But if you are, drinking 4-5 cups is good for you: http://www.nejm.org/doi/full/10.1056/NEJMoa1112010#t=article.

## Other Evidence Roundups

* [Sarah Constantin's *Life Extension Possibilities*](https://srconstantin.wordpress.com/2017/01/24/life-extension-possibilities/)
* [Which Risks of Dying are Worth Taking?](http://www.spencergreenberg.com/2013/10/which-risks-of-dying-are-worth-taking/#.WN64JlwrLCI)

## Methodological Topics

* [Micromort](http://en.wikipedia.org/wiki/Micromort "Wikipedia: Micromort")
* [Microlife](https://en.wikipedia.org/wiki/Microlife)

```{r}
save.image("../data-cache/Longevity.RData")
```

# References
