---
title: "Vitamin D"
bibliography: Longevity.bib
---

```{r, message=FALSE}
library("htmltab")
library("magrittr")
library("plyr")
library("tidyverse")
library("plotly")
library("intubate")
library("knitr")
library("metafor")
library("BayesianFirstAid")
library("bayesmeta")
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
load('../data-cache/Longevity.RData')
```

```{r, results='asis'}
source("infobox.R")
infobox(list(
  "Status" = "Incomplete",
  "Epistemic Status" = "Speculative",
  "Fork of" = tags$a(href='http://www.gwern.net/Longevity', "Gwern's Longevity"),
  "First Revision" = "2016-10-14",
  "Last Revision" = as.Date(file.info("drafts/Longevity.Rmd")$mtime)
))
```

A review of vitamin D meta-analyses reporting on all-cause mortality yields a meta-analytic result of $RR\approx0.96$ with little heterogeneity or signs of bias. Reproducing it as a Bayesian random-effects meta-analysis gives an 84% posterior predictive probability that $RR<1.0$. The expected life expectancy gain is 0.33 years or \$16,800, while total cost of vitamin D supplementation is estimated at \$761, for a profit of \$15,300. The probability of being profitable is estimated to be 83%. The optimal age at which to begin supplementation is 24 years old.

## Background

Vitamins D2 and D3 have been researched intensively ever since their discovery for possible benefits for a range of problems, starting with rickets and expanding from there to everything under the sun. Interest in vitamin D is sparked by many correlational results showing low levels of vitamin D in modern Western populations (unsurprising, given how little time people spend in the sun compared to historically) and how low levels correlate and predict with all sorts of bad things - particularly all-cause mortality.

The downside is that the literature is so large and heterogeneous that it's hard to deal with and for many of the uses investigated, there is conflicting evidence. We are fortunate, however: the large research area of bone fractures/[falls](http://en.wikipedia.org/wiki/Falls%20in%20older%20adults "Wikipedia: Falls in older adults")/osteoporosis in the elderly (collectively, leading causes of death & disability) means many (relatively) large RCTs in (relatively) healthy people are available with meaningful death rates reported. For particularly comprehensive reviews, see [Newberry et al 2014](https://www.ncbi.nlm.nih.gov/books/NBK253540/ "Vitamin D and Calcium: A Systematic Review of Health Outcomes (Update)") and [Theodoratou et al 2014](http://www.bmj.com/content/348/bmj.g2035.full "Vitamin D and multiple health outcomes: umbrella review of systematic reviews and meta-analyses of observational studies and randomised trials").

In addition, vitamin D3 also has considerable plausibility for being profitable: it can be taken in extremely large doses (>10-100x normal daily doses) without much sign of toxicity (leading to a fair number of trials which ensure compliance by just injecting mega-doses of vitamin D every month or year), [hypervitaminosis D](http://en.wikipedia.org/wiki/hypervitaminosis%20D) is rare and not usually fatal, side-effects are rare and minimal, it is already taken by many millions of healthy and sick people without noticeable problems, and vitamin D costs ~\$15/year.

## Causality

### Meta Analyses

The overall thrust of the meta-analyses is that there is consistent and substantial evidence that regular vitamin D3 supplementation reduces all-cause mortality somewhere in the RR=0.90-1 range, that vitamin D2 may be worse but calcium seems irrelevant, and this effect seems to be general: the observable between-study heterogeneity is very small despite large differences in subject populations by gender, age, and dosage, and subgroup analyses do not find the effect confined to particular demographics or that a reduction in a particular disease is driving the ACM reduction (which while potentially due to lack of power, would be consistent with the generality of benefits in both the correlational and Mendelian randomization studies).

The most comprehensive meta-analysis by my count is Bolland et al 2014, which pools _k_=41 to get an estimate of RR 0.96 (0.93-1.00; _p_=0.04) based on an experimental death rate of 3824/40379=0.0947 vs control death rate of 3950/40794=0.0968. The meta-analysis itself can be reproduced given Bolland's forest plot & table, Figure 5.

```{r}
vitaminD <- data.frame(
  "Study"= c("Inkovaara", "Corless", "Ooms", "Lips A", "Komulainen", "Meyer", "Bischoff", "Cooper", "Latham", "Trivedi", "Avenell", "Harwood", "Aloia", "Flicker", "Grant", "Broe", "Burleigh", "Lappe", "Lyons", "Smith", "Björkman", "Chel", "Prince", "Zhu", "Lips B", "Sanders", "Glendenning", "Inkovaara", "Chapuy A", "Dawson-Hughes", "Baeksgaard", "Krieg", "Chapuy B", "Harwood", "Meier", "Brazier", "Grant", "Porthouse", "WHI trials", "Bolton-Smith", "Zhu", "Salovaara"),
 "Year"=c(1983, 1985, 1995, 1996, 1998, 2002, 2003, 2003, 2003, 2003, 2004, 2004, 2005, 2005, 2005, 2007, 2007, 2007, 2007, 2007, 2008, 2008, 2008, 2008, 2010, 2010, 2012, 1983, 1992, 1997, 1998, 1999, 2002, 2004, 2004, 2005, 2005, 2005, 2006, 2007, 2008, 2010),
  "E.deaths"=c(41, 8, 11, 223, 2, 169, 1, 0, 11, 224, 4, 24, 1, 76, 438, 5, 16, 4, 947, 355, 27, 25, 0, 0, 1, 40, 2, 2, 258, 2, 0, 21, 70, 17, 0, 3, 221, 57, 744, 0, 0, 15),
  "E.n"=c(181, 41, 177, 1291, 232, 569, 62, 93, 121, 1345, 70, 113, 104, 313, 2649, 99, 101, 446, 1725, 4727, 150, 166, 151, 39, 114, 1131, 353, 353, 1634, 187, 80, 124, 389, 75, 30, 95, 1306, 1321, 18176, 62, 39, 1718),
  "C.deaths"=c(26, 8, 21, 251, 2, 163, 4, 1, 3, 247, 3, 5, 2, 85, 460, 2, 13, 18, 953, 354, 9, 33, 1, 2, 0, 47, 0, 0, 274, 2, 1, 26, 43, 5, 1, 1, 217, 68, 807, 1, 2, 13),
  "C.n"=c(146, 41, 171, 1287, 232, 575, 60, 94, 122, 1341, 64, 37, 104, 312, 2643, 25, 104, 734, 1715, 4713, 68, 172, 151, 81, 112, 1125, 333, 333, 1636, 202, 80, 124, 194, 37, 25, 97, 1332, 1993, 18106, 61, 41, 1714),
  "Calcium"=c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1)
)

rma(measure="RR", ai=E.deaths, bi=(E.n-E.deaths), ci=C.deaths, di=(C.n-C.deaths), data=vitaminD, method="REML")
```

Since I^2=0, this is equivalent to a fixed-effect meta-analysis:

```{r}
rma(measure="RR", ai=E.deaths, bi=(E.n-E.deaths), ci=C.deaths, di=(C.n-C.deaths), data=vitaminD, method="FE")
```

(Oddly, the per-study mortality rates in the Bolland forest plot/table don't add up to the claimed totals, but are short by a few dozen each; since the RR calculated by `metafor` works out to be about the same, I assume it has something to do with how Bolland decided to handle including multiple subgroups from studies and is not important.)

Turning from frequentist to Bayesian methods, we could take the fixed-effect at face-value and use a Bayesian proportion test, for example:

```{r}
bayes.prop.test(c(sum(vitaminD$E.deaths), sum(vitaminD$C.deaths)), c(sum(vitaminD$E.n), sum(vitaminD$C.n)))
```

Using BFA has the downside that an I^2=0 doesn't guarantee that there is no heterogeneity (it'd be surprising if there wasn't) and gives an overly narrow predictive distribution, and BFA doesn't easily include our informative priors. We can switch to [bayesmeta](https://cran.r-project.org/web/packages/bayesmeta/index.html) which is easier to use than JAGS since it's built on metafor:

```{r}
brem <- escalc(measure="RR", ai=E.deaths, bi=(E.n-E.deaths), ci=C.deaths, di=(C.n-C.deaths), data=vitaminD) %>%
          bayesmeta(mu.prior.mean=0, mu.prior.sd=0.35^2)
brem
```

Converting from log RRs to RR:

```{r}
exp(c(-0.080152143106, -0.036152849917, 0.008768546338))
```

Posterior predictive distribution of RR values including heterogeneity

```{r rrPP}
rrPP <- exp(brem$rposterior(1000, predict=TRUE)[,2])
```

Probability that vitamin D will reduce all-cause mortality in unobserved populations:

```{r}
sum(rrPP<1) / length(rrPP)
```

## Benefit

Expected life expectancy increase:

```{r}
RR <- mean(rrPP)
years <- mean(sapply(rrPP, lifeExpectancyGain))
years * valueOfDALY
```

Hence, vitamin D supplementation could be worth <=\$16.8k.

## Cost

### Financial

The meta-analyses do not have anything remotely near a consensus about dosage (nor do the correlational studies, despite much ink spilled), so it probably does not matter much. I have in the past purchased 360x5000IU vitamin D3 pills for [\$10.99](http://amazon.com/gp/product/B0050MYHBQ/?tag=gwernnet-20 "Doctor's Best Vitamin D3 5000iu Soft-gels, 360"), for an annual cost of \$11.14\. Taking a single pill takes perhaps a sixth of a minute, which again we value at \$8/hr. So if we began taking at age 30:

```{r}
moneyCost <- (10.99 / 360) * 365
timeCost <- ((1/10)/60) * 365.25 * 8
vitaminDAnnualCost <- moneyCost + timeCost
vitaminDTotalCost <- vitaminDAnnualCost * lifeExpectancy(RR, 30, Inf)
vitaminDTotalCost
```

#### Dosage

To go into some more detail about the dosing issue, ne of the more recent meta-analyses to discuss dose in connection with all-cause mortality, [Autier 2014](http://www.gwern.net/docs/vitamind/2014-autier.pdf), says

> …Results of meta-analyses and pooled analyses consistently showed that supplementation could significantly reduce the risk of all-cause mortality, with relative risks ranging from 0.93 to 0.96 (table 4). Most trials included elderly women and a sizeable proportion of individuals were living in institutions. Decreases in risks of death were not associated with trial duration and baseline 25(OH)D concentration. Mortality reductions in trials that used doses of 10-20 μg [400-800IU] per day of vitamin D seemed greater than were reductions noted with higher doses.(13,14)
> 
> *   13\. Bjelakovic G, Gluud LL, Nikolova D, et al. "Vitamin D supplementation for prevention of mortality in adults". Cochrane Database Syst Rev 2011; 7: CD007470\. [[2014](http://onlinelibrary.wiley.com/doi/10.1002/14651858.CD007469.pub2/pdf) update]
> *   14\. Rejnmark L, Avenell A, Masud T, et al. ["Vitamin D with calcium reduces mortality: patient level pooled analysis of 70,528 patients from eight major vitamin D trials"](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3410276/). J Clin Endocrinol Metab 2012; 97: 2670-81.

(1μg=40IU, so 10μg=400IU, 20μg=800IU, and 1250μg=5000IU.)

I'm not sure I agree. The mechanistic theory and correlations do not predict that 400IU is ideal, it doesn't seem enough to get blood serum levels of 25(OH)D particularly higher, and I don't read Rejnmark the same way: the [Figure 3 forest plot](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3410276/figure/F3/), to me, shows that after correcting for Smith's use of D2 rather than D3 (D2 usually performs worse), that there are too few studies using higher doses to make any kind of claim ([Table 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3410276/table/T1/); almost all the daily studies use <=20μg), and the studies which we do have tend to point to higher being better within this restricted range of dosages.

That said, I cannot prove that 5k IU is equally or more effective, so if anyone is feeling risk-averse or dubious on that score, they should stick with 800IU doses.

A year of daily 800IU doses costs almost the same as higher IU dosages since the vitamin D itself is only a small part of the cost of manufacturing. Alternately, if one is unconcerned about the different between daily and more intermittent doses (reasoning that due to the fat-solubility it should not make any difference), one could take a 5k IU dose on a weekly basis, thereby cutting the annual dose cost from ~\$11 to ~\$2.

### Side-effects

Due to rarity, effectively zero: the listed side-effects are all so rare or minor that I can't come up with any reasonable estimate of cost.

## Cost-benefit

Bringing it all together for a 30yo considering vitamin D:

```{r}
benefit <- lifeExpectancyGain(RR, startingAge=30) * valueOfDALY
cost <- vitaminDTotalCost
benefit - cost
```

### Optimal age

```{r}
ages <- sapply(1:120, function(startingAge) profitByAge(startingAge, RR, annualCost = vitaminDAnnualCost, startCost = 0, probabilityPenalty=1))
optimalAge <- which.max(ages)
```

So the optimal age turns out to be around `r optimalAge`.

### Sensitivity

What RR would wipe out the gains from vitamin D, and how likely is an RR that pessimistic?

```{r}
zeroRR <- costToRR(startingAge=optimalAge, annualCost=vitaminDAnnualCost)
sum(rrPP<zeroRR) / length(rrPP)
```

The window of unprofitability is so narrow that it doesn't much change the probability.

# Links for Future Investigation

* [Scott Alexander Thinks Vitamin D3 Supplementation is Bad](http://slatestarcodex.com/2014/01/25/beware-mass-produced-medical-recommendations/)

# References