---
title: "Some Thoughts on 9-11"
---

```{r}
medianAge <- 36.8
lifeExpectancy <- 70.17
casualties <- 2977
lifeYearsLost <- signif((lifeExpectancy - medianAge) * casualties, 2)
```

On this day in 2001, `r casualties` people were [killed by terrorists](https://en.wikipedia.org/wiki/September_11_attacks). Stop for a moment, and try to feel the size of that number. You are one person. `r casualties` people is `r casualties` times the number of people you are. `r casualties` lives were needlessly cut short. Collectively, those people might have lived another `r formatC(lifeYearsLost, format="d", big.mark=',')` years.

```{r}
dailyGrossMortality <- 1.6e5
```

Now take a deep breath. The following day, roughly `r formatC(dailyGrossMortality, format="d", big.mark=',')` more people died. That's more than `r floor(dailyGrossMortality/casualties)` times the number of people killed by terrorists on 9-11. Try to make yourself feel the size of THAT number. That's vastly more people than you will ever meet in your lifetime. It's more than triple the size of my hometown. Imagine three small towns not too far from where you are right now, wiped off the map.

```{r}
daysSince <- as.numeric(Sys.Date()-as.Date("2001-09-11"))
dead <- signif(daysSince * dailyGrossMortality, 2)
```

Every day since 9-11-2001, [approximately `r formatC(dailyGrossMortality, format="d", big.mark=',')` people have died](https://en.wikipedia.org/wiki/Mortality_rate). There have been `r daysSince` days since then, meaning `r formatC(dead, format="d", big.mark=',')` people have died. That's more than the [entire population of Europe](https://www.google.com/search?q=Population+of+Europe), snuffed out. Try (and fail) to wrap your head around that number. It's just too overwhelming. Our brains don't (and can't) work that way.

Instead, just think for a moment how different the world might be if we were able to prioritize our spending on things that stopped people from dying (like evidence-based public health campaigns and medical research), instead of things that make easy re-election campaigns (like defense spending). Think about all the people who could have been alive right now. It's much bigger than the people who were killed on 9-11 itself. So, sure, [#NeverForget](https://www.facebook.com/hashtag/neverforget), but don't forget the rest of humanity in the process.
